Fused regression for network inference in multiple species

### 2015-11-07 paper1.tex
hi

\begin{abstract}
Learning gene regulatory networks is essential to our understanding of cell behavior and response to external stimuli. Methods for network inference from large datasets have been developed and used for a wide variety of species; however, these approaches consider the problem as independent in each species even though there is known to be conservation of gene regulation even in distantly related species. We introduce a method that allows data to be pooled across species in a principled way based on gene orthology and demonstrate that it improves network inference even when the orthology is incomplete. We further refine this method with an algorithm that attempts to learn the true conserved subnetwork from a larger set of potentially conserved interactions. 
\end{abstract}

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist adequate data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine data from multiple platforms, lest we lose valuable information from previous studies. The problem of inferring related -- but not necessarily identical -- structure from related -- but not identical -- data is ubiquitous in biology. Multi-source network inference has applications for learning networks in multiple related species, for learning networks associated with distinct processes within the same species, and for learning networks based on heterogenous data sources.

Most algorithms for network inference can not use data from multiple sources. This is particularly problematic when studying a lesser used model organism, or a biological process that is not widely studied. Approaches that do utilize data from multiple sources have largely relied on either learning individual networks and combining the solutions, which is unviable when one or more of the data sources is very limited, or on treating the data as if it came from a single source, which is limited to sources with a high degree of overlap (ie species with many genes in common). Moreover, combining the data into a single source assumes that the underlying networks are identical. In reality, the similarity of networks in related species is unknown. A method is needed that can flexibly combine data sources based on their similarity.

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference. The core of both methods is the observation that biological information about the relatedness of two networks can be used to guess which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our first method -- fused ridge -- method uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when . Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously learn the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different).

In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. We make the assumption that if a pair of TFs are orthologous to one another, then their regulatory effects on a pair of orthologous genes should be similar. This orthology information forms the basis of a large number of constraints on the network inference problems, which allow data from one organism to improve network inference on another organism. This very general framework for multi-species network inference can be extended to large numbers of distinct organisms, each contributing data with only partially overlapping sets of genes.

Orthology -- the measure of gene similarity we use -- is based on readily identifiable sequence similarity. Although sequence similarity is a useful predictor of functional similarity, we recognize that some genes have taken on different functions and therefore may have new regulatory interactions. These cases of neofunctionalization can be difficult to verify by simply comparing independently inferred networks; because network inference is typically underconstrained, there could be other plausible networks in which the coefficients being compared are similar. This is in contrast with our multi-source method: because the constraints we impose favor weight configurations in which fused interaction weights are similar to one another, a solution in which a priori similar regulatory interactions are found to be different is strong evidence that they really are different. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with these evolutionary constraints. When a gene's regulators have changed through evolution, there may be no way of achieving both these goals. Similarly, if a gene has different regulators during different processes, the expression data may conflict with the fusion constraints. In these cases, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

We approach this problem by introducing a saturating penalty function based on work in statistics to develop unbiased regularization penalties for regression. The main difference between the L2 fusion approach and this adaptive fusion approach occurs when the difference between presumed analogous interactions is large despite the fusion penalty: we assume that large differences represent evidence that functions have diverged and represent this, in adaptive fusion, with a relaxation of the fusion penalty. The resulting cost function is non-convex and difficult to optimize; however, we can approximate its solution and obtain deeper insights into functional similarity than is available through orthology or the comparison of separately learned networks.

We test the ability of fused L2 and adaptive fusion to improve network recovery on both synthetic data and Bacillus subtilis and Bacillus anthracis. We explore the circumstances under which each approach is optimal, and evaluate the robustness of adaptive fusion to incorrect orthology, simulating the biologically relevant case of neofunctionalization.

\section{Methods}
\subsection{Gene regulatory network}
We represent gene regulatory networks as a matrix of weights, where rows index transcription factors and columns index genes. The weight in a given position represents the regulatory weight of that TF on that gene; positive weights represent activation, negative weights represent repression, and 0 weights represent the absence of an interaction. We begin with an existing framework for estimating the interaction weights in a gene regulatory network (GRN). In the Inferelator algorithm, linear differential equations are used to model changes in gene expression. The rate at which the observed mRNA expression of gene $i$, $x_i$, changes is governed by degradation of existing transcripts with rate $\alpha$ plus a linear combination of transcription factor (TF) expressions. 
\begin{equation}
\frac{\mathrm d}{\mathrm d t} x_i = -\alpha_{i}x_{i} + \sum \beta_{i,j}x_{j}
\end{equation}
We are interested in learning $\beta$, the matrix defining the influence of each TF on each gene. We fix the decay rate $\alpha$ for all genes, and set it assuming a time-constant of 10 minutes [CITATIONS], as in [CITATIONS]. Let $x_i(t)$ be the expression of gene $i$ at time $t$. Given time-series data on the expression of gene $i$ at timepoints $t_k$ and $t_{k+1}$, we can approximate the rate of change of $x_i$ as $x_i'(t_k)=\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k}$. We can treat steady-state data as having a derivative of zero. This gives us, for each gene $i$ and time $t_{k}$ an equation

\begin{equation}
\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k} + \alpha_{i}x_{i}(t_k)= \sum \beta_{i,j}x_{j}(t_k)
\end{equation}
\noindent We can summarize these equations in matrix form as
\begin{equation}
Y = X \beta 
\end{equation}
\noindent We approach the learning of the $\beta$ matrix using regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is essentially the same basic formulation as was used in the Inferelator algorithm, which we extend to the case of simultaneously inferring the GRNs of multiple species.

@@

When there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism), the choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). Part of the appeal of the approach, however, is that it allows pooling of data even when there is incomplete orthology. By introducing constraints on the similarity of individual interactions, rather than on the networks as a whole [CITE IBM?], we can pool some information across species even when an arbitrarily small fraction of genes have orthologs. This is particularly useful when dealing with a large number of species; pairwise orthologies may be nearly complete even when the number of genes present in every organism is small.

\subsection{Solving fused L2 problems using augmented matrices}
The problem of solving a multi-valued regression problem can be converted into the problem of solving a single valued regression problem by vectorization.
Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
$y =$ vertically concatenate the columns of $Y$

$$\underset{w}{min} ||Aw - y||_2^{~2}$$

We can then define $W_{i,j} = \beta_{(j-1) \times N + i, (j-1) \times M + j}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication

@@

\begin{figure}[h!]
    \centering
    \includegraphics[scale=.5,trim=0mm 0mm 0mm 0mm,clip]{direct_solve.png}
\end{figure}

@@

where the $j$s refer to transcription factors. In subproblem 1, $j_1$ regulates gene $g$; $j_1$ is orthologous to $j_2$, a transcription factor of subproblem 2, which regulates $h$, a gene which is orthologous to gene $g$. There exists a chain because ($j_1$, $g$) is a TF, gene pair which is fused with ($j_2$, $h$), which is fused to ($j_3$, $k$).
We use depth-first search to identify linked columns of $Y$, and then solve the fused regression problem directly by vectorizing those columns.
In order to incorporate the fusion penalty, we append to the resulting design matrix one row \emph{per connected constraint}, which contains $\lambda_S$ in the column associated with $Y_g,i,j,$ and $-\lambda_S$ to the column associated with $Y_h, i',j'$ for constraint $(Y_g,i,j,Y_h,i',j')$. Call this resulting matrix $X_P$.
The vector $Y$ is augmented by appending an appropriate number of zeros to produce $Y_P$. As a result, each entry in the output of $X_P b$ in the columns associated with each constraint contain $b_{i,j} - b_{i',j'}$.
These entries each contribute  $||b_{i,j} - b_{i',j'}||^{~2}$ to the squared error $||X_Pb - Y_P||_2^{~2}$

@@

With a saturating fusion penalty, fused interactions which appear to be very different based on expression data are allowed to unfuse from one another. A closely related problem has been studied in the context of LASSO regularization, where it was shown by Fan and Li that using a saturating penalty retains many of LASSO's desireable properties while removing its bias towards 0. They further showed that, although the resulting loss-function is nonconvex, good results can be obtained with a local quadratic approximation of gradient descent. Several saturating penalties, such as SCAD [Fan Li 2001] and MCP [Zhang], have been discussed in the context of sparse regression. We introduce a modified form of MCP to the problem of penalizing differences between fused coefficients. The principle difference between the penalty we adopt and SCAD/MCP is that both of these penalties are L1 like at the origin, producing sparse solutions. Because we are penalizing differences in interaction weights, rather than the weights themselves, there's no particular reason to prefer sparsity.

@@

As in [Fan Li], we use solve using iterative local quadratic approximation.

@@

\subsection{Model selection}
We can define a leave-out set of priors and choose parameters based on cross-validated AUPR on the leave-out set.

@@

Following generation of simulated data, we may introduce error into the orthology mapping. This can take the form of discarding a specified fraction of true orthologies (governed by a false-negative rate), or by introducing random false orthologies (governed by a false-positive rate). For convenience, the false-positive rate is specified in units of the number of true orthologs, and not the number of possible orthologs.

@@

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{l2fusionquick.png}
  \caption{\label{fig:figure1} We generate a series of networks with 20 TFs by 200 genes, with 75\% sparsity, 0.1 measurement noise, and 0.1 fusion noise. We hold lamP and lamR constant at 1.0 and 2.0 respectively. The x-axis represents lamS and the y-axis represents the proportion of nodes which are orthologous. As the networks increase in similarity, performance improves and benefits from a large fusion penalty weight. NOTE: This figure shows AUPR. $R^2$ behaves as expected with respect to $\lambda_S$ (networks with large differences between fused coefficients have best $R^2$ at intermediate $\lambda_S$). This doesn't really seem to be the case with AUPR. Maybe these early figures should be replaced with $R^2$}
  \end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{increase_data2.png}
  \caption{\label{fig:figure1} We fuse two networks, both with 25 TFs by 50 genes, using lamS=0 and nonzero lamS, and hold the number of conditions in network 1 constant while varying the number of data points for network 2. We plot the performance of network 2.}
  \end{center}
\end{figure}

\subsection{Network similarity and fusion performance}
We assessed how varying the similarity of simulated networks affected network inference performance with fusion. The main factors governing similarity between our generated networks were the extent (and accuracy) of the orthology mapping, and the magnitude of variability between analogous interactions. We conducted a series of simulations to assess the effect of varying orthology coverage on networks in which analogous interactions varied little from one another. In order to determine the effect of the size of the orthology mapping on network inference performance, we simulated a series of 20 TFs by 200 genes networks at several values of percent orthology coverage.

As expected, increasing the weight of the fusion constraint $\lambda_S$ improved network recovery (Figure 2). Networks with higher orthology coverage saw a greater benefit from fusion, although the improvement was not limited to analogous interactions. One way of assessing the effect of fusion on network inference performance is to compute an ``exchange-rate'' between conditions in each network. Suppose you are interested in determining a regulatory network for species A, a model organism which is experimentally inconvenient. A related species, B, is easier to collect data for, but of less intrinsic interest. Our simulation results suggest that adding conditions from species B to a fused network inference problem has an effect on network A recovery that is similar to adding conditions from A at a reduced rate (Figure 3).

We also assessed the effect of varying the standard deviation of the distribution of differences between analogous interactions on network performance. FILL IN. MENTION PRIOR INTERPRETATION

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{con_noncon.png}
  \caption{\label{fig:figure4} We looked at four different sets of networks, each with 20 TFs by 200 genes, where each set had different amounts of orthologous genes. We plotted the performance on the whole network, the constrained, and non-constrained portions of the network as we increased the fusion penalty weight.}
  \end{center}
\end{figure}

\subsection{Fused regression improves performance on both the constrained and non-constrained parts of the network}
We wanted to know if using fused regression on networks with only conserved subgraphs affected the recovery of the non constrained parts of the networks (Figure 4). We used sets of two networks with 20 TFs and 200 genes each, where each set had 0\%, 25\%, 50\%, or 75\% of the TFs and genes in orthology groups. We varied lamS and computed AUPR using cross-validation on the portion of the network that had fusion constraints, the portion without any orthology information, and on the whole network. On the fused networks, performance improved as the fusion penalty weight increased; interestingly, performance gains were seen even the portion of the network where no orthology information is known. By constraining part of an underdetermined system, we obtained gains in both parts.

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{plot_betas_scad_l2.png}
  \caption{\label{fig:figure1} We used two networks, each with 20 TFs by 200 genes, 50\% of which are orthologous. We include as many false positive orthologs as true positives, and color the points solved using true orthology information green and the points solved using false orthology information red. We randomly select an interaction from each network, and color this point blue. We solve the networks using fused L2.}
  \end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{plot_betas_scad.png}
  \caption{\label{fig:figure1} We used two networks, each with 20 TFs by 200 genes, 50\% of which are orthologous. We include as many false positive orthologs as true positives, and color the points solved using true orthology information green and the points solved using false orthology information red. We randomly select an interaction from each network, and color this point blue. We solve the networks using adaptive fusion.}
  \end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{scadf.png}
  \caption{\label{fig:figure1} Here we use two 10 TF by 200 gene networks, 50\% of which are orthologous. We include as many false positive orthologs as we have true positives, and apply adaptive fusion to the beta values with lamS=3. We plot the distribution of lamS learned for false orthologs.}
  \end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{scadt.png}
  \caption{\label{fig:figure1} Here we use two 10 TF by 200 gene networks, 50\% of which are orthologous. We include as many false positive orthologs as we have true positives, and apply adaptive fusion to the beta values with lamS=3. We plot the distribution of lamS learned for true orthologs.}
  \end{center}
\end{figure}

\subsection{Adaptive fusion successfully identifies and unfuses 'neofunctionalized' genes}
We recognize that orthology prediction is not necessarily a perfect proxy for functional conservation. We implemented an adaptive fusion algorithm that attempts to optimize a nonconvex saturating penalty function on differences between fused interactions. Pairs of interactions which are very dissimilar even after fusion, which sit in the flat portion of this penalty function, are effectively ``unfused''. We can interpret this ``unfusing'' as evidence for neofunctionalization of one or more of the genes involved in this pair of interactions.

We performed a simulation to assess the ability of our adaptive fusion algorithm to learn which parts of two input networks are conserved when orthology information is faulty, to mimic orthologs which are not functionally analogous  (Figure 5-8). We generated synthetic fused networks and introduced error in the fusion constraints by adding false positives and negatives to the orthology information given to the solver. Because we knew which entries in the orthology mapping were ``fake'' (not reflected in the generation of the networks), we could correctly label fusion constraints that involved one or more ``fake'' mappings. We verified that adaptive-fusion unfused mostly ``fake'' interactions, while leaving truly analogous interactions fused.

\begin{figure}
\begin{center}
  \includegraphics[scale=0.45]{test_scad_opt_params3.png}
  \caption{\label{fig:figure1} We use three sets of 20 TF by 200 gene networks, 50\% of which are orthologs, and solve with orthology information containing 0:1, 0.5:1, and 1:1 false positive to true positive. We plot AUPR under fused L2 and adaptive fusion.}
  \end{center}
\end{figure}

\subsection{Adaptive fusion improves network inference}
We wanted to determine whether adaptive fusion is able to improve learning of GRNs. That is, beyond correctly identifying functional conservation and recovering the performance lost by fusing neofunctionalized genes, is our method able to produce more accurate networks than those learned without fusion? (Figure 9) We again generated synthetic fused networks and added false positive and negative orthologs, then used fused L2 and adaptive fusion approaches to solve the underlying network. Fused L2 incorporated flawed orthology information, resulting in degradation of performance as we increased orthology error. Performance increased using adaptive fusion, pointing to its utility in network inference in addition to identifying neofunctionalization.

\section{Bacterial data}
We used gene-expression data from B. subtilis and B. anthracis in order to assess performance gains of fused regression on real data. Our B subtilis data set consists of 266 time-series and steady-state observations of 4891 genes during development. Our B. anthracis dataset consists of 72 time-series and steady-state observations of 5536 genes comprising data from developmental and iron-starvation conditions. There were 247 known Transcription Factors (TFs) in the B. subtilis dataset, and 248 TFs in the B. anthracis dataset.

We obtained 1870 one-to-one orthologs from Inparanoid, which produced 177650 fusion-constraints between gene interactions within the two species. Although this number seems very large, it represents only $14.7\%$ of the regulatory interaction matrix in B. subtilis and $12.9\%$ in B. anthracis. The use of such dissimilar species is a test of whether fusion can improve network-inference even when the overlap in interactions is very small.

>>>>

\begin{abstract}
Learning gene regulatory networks is essential to our understanding of cell behavior and response to external stimuli. Methods for network inference from large datasets have been developed and used for a wide variety of species; however, most approaches consider the problem as independent in each species even though there is known to be conservation of gene regulation even in distantly related species. We introduce a method that allows data to be pooled across species in a principled way based on gene orthology and demonstrate that it improves network inference even when the orthology is incomplete. We further refine this method with an algorithm that attempts to learn the true conserved subnetwork from a larger set of potentially conserved interactions. We demonstrate the utility of our method not only in cross species network inference, but also in merging data sets and incorporating binding similarity predictions.
\end{abstract}

\section{Introduction}
As the volume and variety of genome scale data have exploded, the goal of accurately modeling gene regulatory networks has become more attainable \cite{bonneau_predictive_2007, ciofani_validated_2012, carro_transcriptional_2010}. Large-scale data collection efforts have contributed to the development of high quality networks, but there remain many biological processes and organisms for which there does not exist adequate data. Furthermore, as new technologies are developed and some old ones are replaced, such as RNAseq and microarray, it becomes important to be able to combine data from multiple platforms, lest we lose valuable information from previous studies. The problem of inferring related -- but not necessarily identical -- structure from related -- but not identical -- data is ubiquitous in biology. Multi-source network inference has applications for learning networks in multiple related species, for learning networks associated with distinct processes within the same species, and for learning networks based on heterogenous data sources.

We present two methods for network inference based on linear estimates of gene expression dynamics, extending existing linear dynamical-systems methods for network inference. The core of both methods is the observation that biological information about the relatedness of two networks can be used to approximate which network coefficients should be similar to one another in a multi-source network inference problem (ie orthologous TFs should regulate orthologous genes), and that these constraints can be efficiently represented as penalties in a least-squares regression problem. Our first method -- fused ridge -- method uses an L2 penalty on the differences between a priori similar interactions, and is useful where the relationships between networks (similarity of genes between data sets) is reliable. In the case where both networks contain an identical set of genes and TFs, this approach can be thought of as parametrically interpolating between treating the data sources separately and combining them together. However, our method allows useful pooling of data even when the overlap between genes is incomplete. Our second method -- adaptive fusion -- uses a non-concave saturating fusion penalty to simultaneously learn the constrained networks and to learn which constraints should be relaxed (ie which parts of the network are genuinely different).

In the case of multiple species, there has been a great deal of work showing that functional conservation exists in gene regulatory networks even across large evolutionary distance. We make the assumption that if a pair of TFs are orthologous to one another, then their regulatory effects on a pair of orthologous genes should be similar. This orthology information forms the basis of a large number of constraints on the network inference problems, which allow data from one organism to improve network inference on another organism. This very general framework for multi-species network inference can be extended to large numbers of distinct organisms, each contributing data with only partially overlapping sets of genes. We use our knowledge about the relationship between nodes across networks in order to constrain the network inference problem; this approach has many applications, from improving network inference of multiple species to combining data sets to leveraging knowledge about similarity in binding sites.

Existing multi species approaches often use orthology as a proxy for functional conservation \cite{penfold_inferring_2015, joshi_multi-species_2015, kashima_simultaneous_2009, zhang2010nearly}, or attempt to learn functional similarity via expression data \cite{gholami_cross-species_2010}. Orthology -- the measure of gene similarity we use -- can be approximated using readily identifiable sequence similarity. Sequence similarity is often a useful predictor of functional similarity \cite{wilson_assessing_2000}. Our fused L2 method minimizes a cost function that strives to simultaneously fit expression data in each species and produce networks that are consistent with evolutionary constraints created using orthology. However, we recognize that some genes may have taken on different functions and therefore may have new regulatory interactions. Gene duplications and neofunctionalization in multi-species network inference \cite{eisen_phylogenomics:_1998}, and changes in chromatin configuration in multi-cell line network inference \cite{li_role_2007}, are cases where enforcing fusion of constraints does not produce networks which reflect true differences in regulation between networks. These cases can be difficult to learn by simply comparing independently inferred networks; because network inference is typically underconstrainted \cite{marbach_revealing_2010-1}, there could be other plausible networks in which the coefficients being compared are similar. Our adaptive fusion method allows us to learn differences in networks predicted to be similar based on orthology or gene matching: because the constraints we impose favor weight configurations in which fused interaction weights are similar to one another, a solution in which a priori similar regulatory interactions are found to be different is strong evidence that they really are different. When differences in regulatory interactions arise where they are predicted to be similar, the network inference procedure would be improved by relaxing or ignoring these evolutionary constraints. However, we don't know a priori which constraints to enforce and which to relax.

We approach this problem by introducing a saturating penalty function based on work in statistics to develop unbiased regularization penalties for regression \cite{zhang2010nearly, fan2001variable}. The main difference between the L2 fusion approach and this adaptive fusion approach occurs when the difference between presumed analogous interactions is large despite the fusion penalty: we assume that large differences represent evidence that gene functions have diverged and represent this, in adaptive fusion, with a relaxation of the fusion penalty. The resulting cost function is non-convex and difficult to optimize; however, we can approximate its solution and obtain deeper insights into functional similarity than is available through orthology or the comparison of separately learned networks.

We test the ability of fused L2 and adaptive fusion to improve network recovery on both synthetic data and Bacillus subtilis and Bacillus anthracis, showing the applicability of our method in combining different datasets, and leveraging similarity across organisms as well as within a network in order to improve network inference. We explore the circumstances under which each approach is optimal, and evaluate the robustness of adaptive fusion to incorrect orthology, simulating the biologically relevant case of neofunctionalization.

\section{Methods}
\subsection{Gene regulatory network}
We represent gene regulatory networks as a matrix of weights, where rows index transcription factors (TFs) and columns index genes. The weight in a given position represents the regulatory weight of that TF on that gene; positive weights represent activation, negative weights represent repression, and 0 weights represent the absence of an interaction. We begin with an existing framework for estimating the interaction weights in a gene regulatory network (GRN). In the Inferelator algorithm, linear differential equations are used to model changes in gene expression \cite{bonneau_inferelator:_2006-1} The rate at which the observed mRNA expression of gene $i$, $x_i$, changes is governed by degradation of existing transcripts with rate $\alpha$ plus a linear combination of transcription factor (TF) expressions. 
\begin{equation}
\frac{\mathrm d}{\mathrm d t} x_i = -\alpha_{i}x_{i} + \sum \beta_{i,j}x_{j}
\end{equation}
We are interested in learning $\beta$, the matrix defining the influence of each TF on each gene. We fix the decay rate $\alpha$ for all genes, and set it assuming a time-constant of 10 minutes \cite{hambraeus_genome-wide_2003, selinger_global_2003}, as in \cite{greenfield_robust_2013}. Let $x_i(t)$ be the expression of gene $i$ at time $t$. Given time-series data on the expression of gene $i$ at timepoints $t_k$ and $t_{k+1}$, we can approximate the rate of change of $x_i$ as $x_i'(t_k)=\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k}$. We can treat steady-state data as having a derivative of zero. This gives us, for each gene $i$ and time $t_{k}$ an equation

\begin{equation}
\frac{x_i(t_{k+1})-x_i(t_k)}{t_{k+1}-t_k} + \alpha_{i}x_{i}(t_k)= \sum \beta_{i,j}x_{j}(t_k)
\end{equation}
\noindent We can summarize these equations in matrix form as
\begin{equation}
Y = X \beta 
\end{equation}
\noindent We approach the learning of the $\beta$ matrix using regression. Because there are typically far fewer conditions than possible regressors (TFs), we introduce a ridge regularization constraint with weight $\lambda_R$ and solve
\begin{equation}
\argmin_\beta\vert \vert X\beta - Y_2 \vert \vert ^2 + \lambda_R \vert \vert \beta \vert \vert ^2
\end{equation}
This is similar to the formulation used in the Inferelator algorithm, which we extend to the case of simultaneously inferring the GRNs of multiple species.

@@

When there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism), the choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). Part of the appeal of the approach, however, is that it allows pooling of data even when there is incomplete orthology. By introducing constraints on the similarity of individual interactions, rather than on the networks as a whole \cite{liu2011temporal}, we can pool some information across species even when an arbitrarily small fraction of genes have orthologs. This is particularly useful when dealing with a large number of species; pairwise orthologies may be nearly complete even when the number of genes present in every organism is small.

\subsection{Solving fused L2 problems using augmented matrices}
The problem of solving a multi-valued regression problem can be converted into the problem of solving a single valued regression problem by vectorization.
Specifically, if $Y \in R^{N \times M}$ then we form $A =$ diagonally concatenate $X$, $M$ times (with the remaining entries 0),
$y =$ vertically concatenate the columns of $Y$. We solve

$$\underset{W}{min} ||AW - y||_2^{~2}$$

%We can then define $W_{i,j} = \beta_{(j-1) \times N + i, (j-1) \times M + j}$ which is the desired form.
This is equivalent to the original problem due to the block structure of matrix multiplication

@@

%\begin{figure}[h!]
%    \centering
%    \includegraphics[scale=.5,trim=0mm 0mm 0mm 0mm,clip]{direct_solve.png}
%\end{figure}

@@

where the $j$s refer to transcription factors. In subproblem 1, $j_1$ regulates gene $g$; $j_1$ is orthologous to $j_2$, a transcription factor of subproblem 2, which regulates $h$, a gene which is orthologous to gene $g$. There exists a chain because ($j_1$, $g$) is a TF, gene pair which is fused with ($j_2$, $h$), which is fused to ($j_3$, $k$).
We use depth-first search to identify linked columns of $Y$, and then solve the fused regression problem directly by vectorizing those columns.
In order to incorporate the fusion penalty, we append to the resulting design matrix one row \emph{per connected constraint}, which contains $\lambda_S$ in the column associated with $Y_g,i,j,$ and $-\lambda_S$ to the column associated with $Y_h, i',j'$ for constraint $(Y_g,i,j,Y_h,i',j')$. Call this resulting matrix $X_P$.
The vector $Y$ is augmented by appending an appropriate number of zeros to produce $Y_P$. As a result, each entry in the output of $X_P b$ in the columns associated with each constraint contain $\lambda_S(b_{i,j} - b_{i',j'})$.
These entries each contribute  $\lambda_S||b_{i,j} - b_{i',j'}||^{2}$ to the squared error $||X_Pb - Y_P||_2^{~2}$

@@

With a saturating fusion penalty, fused interactions which appear to be very different based on expression data are allowed to unfuse from one another. A closely related problem has been studied in the context of LASSO regularization, where it was shown by Fan and Li that using a saturating penalty retains many of LASSO's desireable properties while removing its bias towards 0 \cite{fan2001variable}. They further showed that, although the resulting loss-function is nonconvex, good results can be obtained with a local quadratic approximation of gradient descent. Several saturating penalties, such as SCAD \cite{fan2001variable} and MCP \cite{zhang2010nearly}, have been discussed in the context of sparse regression. We introduce a modified form of MCP to the problem of penalizing differences between fused coefficients. The principle difference between the penalty we adopt and SCAD/MCP is that both of these penalties are L1 like at the origin, producing sparse solutions. Because we are penalizing differences in interaction weights, rather than the weights themselves, there's no particular reason to prefer sparsity.

@@

As in \cite{fan2001variable}, we use solve using iterative local quadratic approximation.

@@

%\subsection{Model selection}
%We can define a leave-out set of priors and choose parameters based on cross-validated AUPR on the leave-out set.

@@

Following generation of simulated data, we may introduce error into the orthology mapping. This can take the form of discarding a specified fraction of true orthologies (governed by a false-negative rate), by introducing random false orthologies (governed by a false-positive rate), or by adding Gaussian noise so that fused interactions are not identical. For convenience, the false-positive rate is specified in units of the number of true orthologs, and not the number of possible orthologs.

@@

\subsection{Network similarity and fusion performance}
Using related data for network inference allows us to borrow statistical power from other sources, effectively increasing the sample size and boosting the sensitivity and specificity of learned interactions. The decision of what other sources to draw from is an important one; at one extreme is combining unrelated data, and at the other is resampling from the same experiments. We simulated this spectrum by creating pairs of networks and varying the similarity between networks. The main factors governing similarity between our generated networks is the extent (and accuracy) of the orthology mapping, and the magnitude of variability between analogous interactions. We conducted a series of simulations to assess the effect of increasing orthology coverage on networks. In order to determine the effect of the size of the orthology mapping on network inference performance, we simulated a series of 20 TFs by 200 genes networks at several values of percent orthology coverage.

As expected, when the conserved subgraph is very similar, increasing the weight of the fusion constraint $\lambda_S$ improved network recovery (Figure 2). Networks with higher orthology coverage saw a greater benefit from fusion, although the improvement was not limited to analogous interactions (Figure 4). When the conservation is weak, and even the portions of the network which are conserved have deviated from each other significantly, performance may not continue to improve with by increasing the fusion weight; we simulate this by adding noise to the conserved subgraph (Figure 3). We show that even for networks where the conserved subgraph is weakly conserved, there exists a 'sweet spot' where fusion regression improves network recovery.

One way of assessing the effect of fusion on network inference performance is to compute an ``exchange-rate'' between conditions in each network. Suppose you are interested in determining a regulatory network for species A, a model organism which is experimentally inconvenient. A related species, B, is easier to collect data for, but of less intrinsic interest. Our simulation results suggest that adding conditions from species B to a fused network inference problem has an effect on network A recovery that is similar to adding conditions from A at a reduced rate.

\subsection{Fused regression improves performance on both the constrained and non-constrained parts of the network}
Our method learns networks simultaneously, using predictions about conservation of TF-gene interactions to share information between sources. Our approach is useful for learning networks from similar sources such as different cell-lines from the same species, where there exists a one-to-one mapping of genes, as well as datasets where the networks are expected to be more different than similar, where there exists limited orthology mapping. For the latter case, we are interested in knowing if performance gains from fused regression are limited to those interactions which have fusion constraints. To test this, we used sets of 20 TF by 200 gene synthetic networks, with varying proportions of orthology information about TFs and genes. We varied lamS and computed AUPR on the portion of the network that had fusion constraints, the portion where TF-gene interactions from one network could be mapped to putative analogous interactions in the second network based off of orthology; the unconstrained subnetwork; and the whole network. As expected, performance improved as the fusion penalty weight increased. Interestingly, performance gains were seen even in the portion of the network that is unconstrained by fusion. By constraining part of an underdetermined system, we obtain gains in even the unconstrained part.

\subsection{Adaptive fusion successfully identifies and unfuses 'neofunctionalized' genes}
We recognize that orthology prediction is not a perfect proxy for functional conservation. We implemented an adaptive fusion algorithm that attempts to optimize a nonconvex saturating penalty function on differences between fused interactions (Figures 5-6). Pairs of interactions which are very dissimilar even after fusion, which sit in the flat portion of this penalty function, are effectively ``unfused''. Because our network inference problem is underdetermined, meaning that multiple networks would fit our data equally well, the ``unfusing'' of constraints is convincing evidence for the rejection of a network which uses orthology prediction to predict similarity in interaction weights. ''Unfusing'' suggests neofunctionalization of one or more of the genes involved in this pair of interactions.

We performed a simulation to assess the ability of our adaptive fusion algorithm to learn which parts of two input networks are conserved when orthology information is faulty, to mimic orthologs which are not functionally analogous  (Figure 7-8). We generated synthetic fused networks and introduced error in the fusion constraints by adding false positives and negatives to the orthology information given to the solver. Because we knew which entries in the orthology mapping were ``fake'' (not reflected in the generation of the networks), we could correctly label fusion constraints that involved one or more ``fake'' mappings. We verified that adaptive-fusion unfused mostly ``fake'' interactions, while leaving truly analogous interactions fused. We compared the recovery of interaction weights which were accurately fused, and recovery of interaction weights which were inaccurately fused due to incorrect orthology information. Because fused L2 heavily penalizes large differences between weights which are predicted to be similar (Figures 8-9), the error for inaccurately fused interactions is higher than solving without fusion. When we use our adaptive fusion approach, we preserve the gain in performance on accurately fused interactions, while reducing the error for weights of interactions involving "fake" orthologies.

\subsection{Cross-species network inference using bacterial data}
We used gene-expression data from B. subtilis and B. anthracis in order to assess performance gains of fused regression on real data. Our B subtilis data set consists of 360 time-series and steady-state observations of 4891 genes during development. Our B. anthracis dataset consists of 72 time-series and steady-state observations of 5536 genes comprising data from developmental and iron-starvation conditions. There were 247 known Transcription Factors (TFs) in the B. subtilis dataset, and 248 TFs in the B. anthracis dataset.

We obtained 1870 one-to-one orthologs from Inparanoid \cite{ostlund_inparanoid_2010}, which produced 177650 fusion-constraints between gene interactions within the two species. Although this number seems very large, it represents only $14.7\%$ of the regulatory interaction matrix in B. subtilis and $12.9\%$ in B. anthracis. The use of such dissimilar species is a test of whether fusion can improve network-inference even when the overlap in interactions is very small.
